Two applications of logic programming to Coq

### 2020-11-04 papers/types2020/sec3.tex
updated todo after chat

\subsection{Focused proof systems with certificates}
\label{ssec:focused}
\todo{Change title? }
\todo{Use same names in Fig and 3: A for Atom, L for ListArgs etc.}

@@

% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
\begin{footnotesize}
\begin{figure}[t]
  %
  \renewcommand{\Gamma}{}
    \newcommand{\bc}[5]{#1#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi\Gamma\vdash A : s}
{(A,s) \in \mathcal{A}
}
% \qquad
% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
\qquad
\infer{\XXi_1\Gamma\vdash k\ l : A}
      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\cup\Gamma\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ ::\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

@@

Since we are only dealing with types that correspond to Horn clauses, and
atomic types are inductively defined, we do not need a $\forall$ rule on the right, although the proof theory would gladly allow it. This means, at the term level, that we
do not have variables in our grammar of terms. Terms are always
applied to a (possibly empty) list of arguments, and as is usual in Pure Type
Systems we have the same grammar for terms and types:
%, as shown in figure~\ref{fig:terms}.
\todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
%\begin{figure}
\begin{align*}
t,u ::= & k\ l \;|\; \forall x : t.\, u \\
l ::= & [] \;|\; (t\ ::\ l)
\end{align*}
% \caption{Grammar of terms}
% \label{fig:terms}
% \end{figure}

\todo{This does not see quite right: D vs A, consider say even}
The proof system is parameterized by  external set of inductive definitions available in
the (implicit) global context: we denote by $\Pscr_A$ the set of type declarations for the
constructors of type $A$, of the form $k : D$ where $D$ is of the form
$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
by $\mathcal{A}$ the set of arity declaration for the inductive types.

@@

Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
terms to terms, namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
when, in the implementation of the product-left rule, we
apply $B$ to the variable \lsti{Tm},
we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In this sense, our calculus uses \emph{implicit} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free. 
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
\todo{Now safe-dest-app is hidden inside \texttt{definition}. Put the rest of the code in the appendix?}
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.

@@

\begin{metanote}
Keep this? Perhaps explain Ev = Term?
\end{metanote}

@@

\begin{lstlisting}
Elpi Tactic dprolog.
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _  Ev G _] _ :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
/* Other clauses for different fpc omitted */
}}.
\end{lstlisting}
The following example show how we can use the above tactic to do
FPC-driven logic programming in Coq and return a Coq proof-term:
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog height 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
\end{lstlisting}

>>>>

\subsection{Proof systems with certificates}
\label{ssec:focused}
\todo{Change title? }
\todo{Use same names in Fig and 3: A for Atom, L for ListArgs etc.}

@@

% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
\begin{footnotesize}
\begin{figure}[t]
  %
  \renewcommand{\Gamma}{}
    \newcommand{\bc}[5]{#1#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi\Gamma\vdash A : s}
{%(A,s) \in \mathcal{A}
  \mbox{call to Coq?}
}
% \qquad
% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
\qquad
\infer{\XXi_1\Gamma\vdash k\ l : A}
      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ ::\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

@@

Since we are only dealing with types that correspond to Horn clauses, and
atomic types are inductively defined, we do not need a $\forall$ rule on the right, although the proof theory would gladly allow it. This means, at the term level, that we
do not have variables in our grammar of terms. Terms are always
applied to a (possibly empty) list of arguments% , and as is usual in Pure Type
% Systems we have the same grammar for terms and types
.
%, as shown in figure~\ref{fig:terms}.
% \todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
% %\begin{figure}
% \begin{align*}
% t,u ::= & k\ l \;|\; \forall x : t.\, u \\
% l ::= & [] \;|\; (t\ ::\ l)
% \end{align*}
% % \caption{Grammar of terms}
% % \label{fig:terms}
% % \end{figure}

\todo{Lookup Coq rules for unfold rule}
The proof system is parameterized by  external set of inductive definitions available in
the (implicit) global context: we denote by $\Pscr_A$ the set of type declarations for the
constructors of type $A$, of the form $k : D$ where $D$ is of the form
$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
by $\mathcal{A}$ the set of arity declaration for the inductive types.

@@

Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
terms to terms, namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
when, in the implementation of the product-left rule, we
apply $B$ to the variable \lsti{Tm},
we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In this sense, our calculus uses \emph{implicit} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free. 
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
\todo{Reput safe-dest-app in code, a bit more details about the unfold case}
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.

@@

\todo{explain Ev = Term as an Elpi thing}

@@

\begin{lstlisting}
Elpi Tactic dprolog.
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _  Ev G _] _ :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
(* Other clauses for different fpc omitted *)
}}.
\end{lstlisting}
The following example show how we can use the above tactic to do
FPC-driven logic programming in Coq and return a Coq proof-term:
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog height 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
\end{lstlisting}
